---
title: "Blockmodeling Tutorial"
author: "Mark Hoffman"
output: html_document
---

# Blockmodeling

Blockmodeling is not implemented in igraph. However, I included some code which will allow you to do it easily. You will need to install a package called sna for it to run. 

To begin, we load igraph.

```{r}
# install.packages("sna")
# install.packages("NetCluster")
# install.packages("corrplot")
# install.packages("data.table")

library(igraph)
```

Next we load in all three relationships for the Florentine family data from my website.

```{r}
marriage_edgelist <- read.csv("http://www.markanthonyhoffman.com/downloads/florentine_marriage_edgelist.csv", stringsAsFactors = FALSE, row.names = 1)

credit_edgelist <- read.csv("http://www.markanthonyhoffman.com/downloads/florentine_credit_edgelist.csv", stringsAsFactors = FALSE, row.names = 1)

partner_edgelist <- read.csv("http://www.markanthonyhoffman.com/downloads/florentine_partner_edgelist.csv", stringsAsFactors = FALSE, row.names = 1)

florentine_attributes <- read.csv("http://www.markanthonyhoffman.com/downloads/florentine_attributes.csv", stringsAsFactors = FALSE)
```

There are three relations: marriage, credit provision and partnership. The first step in blockmodeling is to "stack" the relations, which means to concatenate them into a single, rectangular matrix.

We put them in a list (which is a collection of objects).
```{r}
florentine = list(Credit = credit_edgelist, Partner = partner_edgelist, Marriage = marriage_edgelist)
```


We them use lapply to apply the same function to each relation in the list. We convert each to a network and then get the adjacency matrix of the networks.
```{r}
florentine <- lapply(florentine, FUN = function(x) simplify(graph_from_data_frame(d = x, directed = F, vertices = florentine_attributes)))

florentine = lapply(florentine, FUN = function(x) as.data.frame(as.matrix(get.adjacency(x))))
```

Finally, we use do.call("rbind", x) to stackk all of the matrices by row. 
```{r}
library(data.table)
florentine_stacked = do.call("rbind", florentine)
```

Next, we measure structural equivalence using correlation (that is, we measure the similarity between people in terms of the pattern of their relations). The cor function measures correlation.
```{r}
florentine_cor = cor(florentine_stacked)
```

We can plot the resulting correlation matrix using the corrplot function.
```{r}
library(corrplot)
corrplot(florentine_cor, order = "hclust", tl.col = "black", tl.cex = .6)
```

We can convert similarity to dissimilarity by subtracting it from 1. We convert it to a distance matrix with as.dist().
```{r}
dissimilarity <- 1 - florentine_cor
dissimilarity <- as.dist(dissimilarity)
```

We can use the NetCluster package to cluster the matrix, permute the matrix, and blockmodel. 
```{r}
library(NetCluster)
```

We use the hclust function to cluster the dissimilarity matrix
```{r}
result = hclust(dissimilarity)
n = nrow(florentine_cor)
```

To partition the matrix , we have to cut the resulting hierarchical clustering tree. 
```{r}
num_clusters = 7
clusters <- cutree(result, k = num_clusters)
cluster_cor_mat <- clusterCorr (florentine_cor, clusters)
```

With those clustering results, we split the original matrix stack back into its constituent matrices.
```{r}
florentine_split = split(florentine_stacked, c(rep(1, 116), rep(2, 116), rep(3, 116)))
florentine_split = lapply(florentine_split, as.matrix)
```

And we run the blockmodel function inputting those clustering results.
```{r}
all_output = blockmodel(florentine_split, clusters, glabels = names(florentine), plabels = colnames(florentine_split[[1]]))
```

Finally, we plot the result!
```{r}
plot(all_output)
```