---
title: "Blockmodeling Tutorial"
author: "Mark Hoffman"
output: html_document
---

# Blockmodeling

Blockmodeling is not implemented in igraph. However, I included some code which will allow you to do it easily. You will need to install a package called sna for it to run. 

To begin, we load igraph.

```{r}
# install.packages("sna")
library(igraph)
```

Next load in the bank_wiring.rda data included in the Data folder, which comes from Breiger et. al's 1975 paper on blockmodeling. .rda is a type of R-specific data which can be loaded with the load function. The only thing you need to change is the file path. 

```{r}
load("/Users/markhoffman/gh_repos/sna_teaching/Data/bank_wiring.rda")
```

We can look at the data.  It is a list of matrices. 
```{r}
bank_wiring
```

I made the CONCOR/blocking function work for either matrices or igraph objects, so you could also put a series of igraph objects in a list to perform blockmodeling. Just to show you that it works, let's convert all of the matrices in the bank_wiring list to igrpah objects using the lapply function.
```{r}
bank_wiring <- lapply(bank_wiring, FUN = function(x) graph_from_adjacency_matrix(as.matrix(x)))

bank_wiring
```

Finally, let's source the file which includes the necessary function and run the concor_blockmodel function.
```{r}
source("/Users/markhoffman/gh_repos/sna_teaching/Useful R Functions/blockmodeling and concor.R")
bank_blockmodel <- concor_blockmodel(bank_wiring, p = 2)
```

