---
title: "Ego-centric Network Analysis"
author: "Mark Hoffman"
output: html_document
---

In this tutorial, we will cover basic network analysis in R. We will make our foray into network statistics and measures by analyzing a series of smaller networks (N < 100). Though larger networks are sexier, for pedagogical purposes, it is better to start small, so that we can visually comprehend how the various measures work.

We begin by looking at ego networks using the GSS network module from 2004, as mentioned by Peter in his lecture. We will use the GSS data to become acquainted with measures of network density and heterogeneity.

We then move to exploring transitivity at the macro and micro levels by returning to Countryside High School. 

# Ego Networks

We begin by reading in the data from the GSS network module, which I have included in the "Data" section of the materials for this class.  

```{r}
library(igraph)

gss <- read.csv("Data/gss_local_nets.csv", stringsAsFactors = TRUE) 
gss_some_given <- subset(gss, numgiven > 1)
```

Let's have a look at the data.  You can either click on it in your environment, type View(gss) or:

```{r}
head(gss)
```

There are 42 variables.  The first five concern the attributes of a given respondent: their sex, age, race, partyid and religion. The next 36 make up the "network" part of the GSS Network Module. The structure can be a bit confusing, especially if you haven't read any papers that use this data. The basic idea of the module was to ask people about up to five others with whom they discussed "important matters" in the past six months.  The respondents reported the number of people whom they discussed "important matters": which is the variable "numgiven" in our dataset. They were also asked to detail the relations between those five people: whether they were especially close, knew each other, or were total strangers. This accords to the close variables in the dataset, where, for example, close12 is the closeness of person 1 to person 2, for each respondent. Finally they were asked about the attributes of each of the up to five people in their ego network (sex, race, age).

To see why these are called ego networks, let's take a respondent and graph the relations of the up to five people they said they discussed "important matters" with.  To do so, we have to first turn the variables close12 through close45 into an edge list. 

This requires a tricky bit of code.
First let's just select respondent 3.

```{r}
respondent <- gss[3,]
```

Next, let's make a general edgelist, which can hold the values for the relations between persons 1 and 2, persons 1 and 3 all the way to persons 4 and 5. We can use the combn function to do this - it will give you all of the possible unique, unordered combinations of a vector of size m. Here we do all of the unique combinations of numbers one through five of size 2. 

```{r}
respondent_edge_list <- as.data.frame(t(combn(1:5, 2)), stringsAsFactors = F)
```

Let's change the column names...
```{r}
colnames(respondent_edge_list) <- c("PersonA", "PersonB")
```
 
Next we need to assign the values from the respondent we saved above to a new variable in our edge list. In this next line of code, we assign to the "weight variable" the values of the variables from close12 to close45. 
```{r} 
respondent_edge_list$weight <- as.numeric(respondent[,which(colnames(respondent) == "close12"):which(colnames(respondent) == "close45")])
```

Finally, let's drop all of the ties that are equal to either NA or 0. 
```{r}
respondent_edge_list <- subset(respondent_edge_list, weight == 1 | weight == 2)
```

How does it look?

```{r}
head(respondent_edge_list)
```

Great! We can use this edge list to creat a network, in the same way that we did in the last tutorial. 
```{r}
ego_net <- graph.edgelist(as.matrix(respondent_edge_list[,1:2]), directed = F)
V(ego_net)$name <- 1:vcount(ego_net)

plot(ego_net, vertex.size = 30, vertex.label.color = "black", vertex.label.cex = 1)
```

Now let's add attributes to the nodes of the network. First we build an attribute list.
```{r}
attributes <- data.frame(name = 1:5, sex = NA, race = NA, age = NA, stringsAsFactors = F)

attributes$race =  as.numeric(respondent[,which(colnames(respondent) == "race1"):which(colnames(respondent) == "race5")])

attributes$sex =  as.numeric(respondent[,which(colnames(respondent) == "sex1"):which(colnames(respondent) == "sex5")])

attributes$age =  as.numeric(respondent[,which(colnames(respondent) == "age1"):which(colnames(respondent) == "age5")])

attributes <- na.omit(attributes)
```

And use graph from data frame to build a network with the edge.list and attributes
```{r}
ego_net <- graph_from_data_frame(respondent_edge_list, vertices = attributes)
```

Let's see the structure of the graph with the nodes colored by sex.
```{r}
V(ego_net)$color <- ifelse(V(ego_net)$sex == 1, "blue", "pink")

plot(ego_net, vertex.size = 30, vertex.label.color = "black", vertex.label.cex = 1)
```

# Calculating Network Size and Density

Network size is the number of nodes in a network. To find this, we use the vcount() function. We can also find the number of edges using ecount()
```{r}
vcount(ego_net)
ecount(ego_net)
```

Density captures how many edges there are in a network divided by the total possible number of edges. In an undirected network of size N, there will be  (N * (N-1))/2  possible edges. 

We could calculate this on our own as follows.
```{r}
ecount(ego_net)/((vcount(ego_net) * (vcount(ego_net) - 1))/2)
```

igraph has its own function - graph.density
```{r}
graph.density(ego_net)
```

# Computing Heterogeneity for Continuous variables using Standard Deviation

For continuous variables, we can use standard deviation to measure heterogeneity. 

The sd function returns the standard deviation.
```{r}
sd(V(ego_net)$age, na.rm = T)
```

# Computing Heterogeneity for Categorical Variables using Index of Qualitative Variation (IQV)

The IQV returns the amount of heterogeneity for a categorical variable.

> **IQV**
> IQV = K(100^2 - SUM(Pct^2))/100^2(K-1)

For the first time in this class, we will define our own function, so that we can run IQV, since it isn't part of igraph.
```{r}
IQV <- function(tbl, k) {
  perc <- tbl * 100
  sumsquared <- sum(perc^2)
  sumsquared <- 10000-sumsquared
  heterogeneity <- (k * sumsquared)/((k-1)*(10000))
  if (is.nan(heterogeneity) == TRUE) heterogeneity = NA
  return(heterogeneity)
}
```

prop.table returns the proportion of each category that makes up your dataset. For example, for the respondent we are working with it returns .5 male and .5 female
```{r}
prop.sex <- prop.table(table(V(ego_net)$sex))
IQV(prop.sex, 2)
prop.race <- prop.table(table(V(ego_net)$race))
IQV(prop.race, length(unique(gss$race)))
```

# Advanced Topic: Working with thousands of networks at once

The GSS data we loaded in contains more than 1000 ego networks.  What if we wanted to analyze all of them without having to type in each and every command? We would use something called a loop. Loops automate the repition of actions. Here we will create a function which turns builds an ego net for a given respondent in the GSS dataset and we will use a loop to apply that function to every respondent.

This is how you define a function in R. You give it a name, here "construct_ego_net", and then assign to it the function() function with some arguments: a response from the gss

```{r}
construct_ego_net <- function(respondent){
  
  require(igraph)
  
  # This constructs the edgelist
  respondent_edge_list <- as.data.frame(t(combn(1:5, 2)), stringsAsFactors = F)
  colnames(respondent_edge_list) <- c("PersonA", "PersonB")
  
  # This puts in the values from the response into the edge list
  respondent_edge_list$weight <- as.numeric(respondent[,which(colnames(respondent) == "close12"):which(colnames(respondent) == "close45")])
  
  # We drop unobserved edges
  respondent_edge_list <- subset(respondent_edge_list, weight == 1 | weight == 2)

  # Create an attribute dataframe
  attributes <- data.frame(name = 1:5, sex = NA, race = NA, age = NA, stringsAsFactors = F)
  attributes$race =  as.numeric(respondent[,which(colnames(respondent) == "race1"):which(colnames(respondent) == "race5")])
  attributes$sex =  as.numeric(respondent[,which(colnames(respondent) == "sex1"):which(colnames(respondent) == "sex5")])
  attributes$age =  as.numeric(respondent[,which(colnames(respondent) == "age1"):which(colnames(respondent) == "age5")])
  
  ind <- apply(attributes[,2:4], 1, function(x) all(is.na(x)))
  attributes <- attributes[ !ind, ]
  
  # Graph from data frame
 ego_net <- graph_from_data_frame(respondent_edge_list, vertices = attributes)
  
  return(ego_net)
}
```

Now let's put the function into a loop. First we create a list to save all of the networks that the function will create. Here we use the vector function to create a vector of type list which is the same size as the number of respondents (or rows) in the GSS data. 
```{r}
all_gss_nets <- vector("list", nrow(gss_some_given))
```

Finally we create a loop which runs the above function for each row in the data and saves the results to the list. 

```{r}
for(i in 1:nrow(gss_some_given)){
  all_gss_nets[[i]] = construct_ego_net(gss_some_given[i,])
}
```

Once we have a list of networks, we can use the lapply function to apply the same function to every network in the list. Here is an example with density
```{r}
network_densities = lapply(all_gss_nets, FUN = function(x) graph.density(x))
# Lapply returns a list, so we use unlist to turn the list into a vector
network_densities = unlist(network_densities)
# And we take the mean!
average_density = mean(network_densities, na.rm = T)
```

You could do the same thing for any of the other functions we used above.