---
title: "group_detection"
author: "Mark Hoffman"
date: "5/3/2017"
output: html_document
---

# Group detection

We often just want to find distinct groups of people. We might not require that everyone in a group is connected - it is too stringent. Instead, we could define groups as sets of nodes who have a higher proportion of edges going inwards than outwards - that is, solidarity is strongest within the group.

There are a plethora of group detection algorithms in igraph. I will cover some of them. The full list is: edge.betweenness.community, fastgreedy.community, label.propagation.community, leading.eigenvector.community, multilevel.community, optimal.community, spinglass.community, and walktrap.community.

This webpage has a summary of their pros and cons for an older version of igraph: http://bommaritollc.com/2012/06/summary-community-detection-algorithms-igraph-0-6/

Essentially you should tailor your algorithm choice to your network. Certain algorithms are designed for directed or undirected graphs, and work better with small or large graphs. 
Each algorithm as its own igraph function. These functions produce lists with information about the algorithm results. Element 1 holds a vector that details which group each node is in, which I will refer to as the membership vector. Element 6 holds the modularity of the network given the group detection algorithm. 

For undirected graphs, you can use the optimal or multilevel algorithms.

First we load our data
```{r}
florentine_edgelist <- read.csv("http://www.markanthonyhoffman.com/downloads/florentine_edgelist.csv", stringsAsFactors = FALSE, row.names = 1)

florentine_attributes <- read.csv("http://www.markanthonyhoffman.com/downloads/florentine_attributes.csv", stringsAsFactors = FALSE)

marriageNet <- graph_from_data_frame(d = florentine_edgelist, directed = F, vertices = florentine_attributes)
marriageNet = simplify(marriageNet)
```

```{r}
communityMulti <- multilevel.community(marriageNet)
```

For directed graphs, edge betweenness is generally your best bet, though the walktrap algorithm performs well too.

We might wish to visualize the group structure of networks to see how the algorithms divide the network. To do so, we have to first extract the membership vector and put it in a format the plot function understands.

```{r}
communityMultiGroup <- communities(communityMulti)
```

We can then use the mark.groups argument in plot to see the groups. 
```{r}
plot(marriageNet, vertex.size = 3, vertex.label = NA, mark.groups = communityMultiGroup)
```

Alternatively, we can just color the nodes by group membership.  In this case, we can just use the membership vector.

```{r}
V(marriageNet)$color <- membership(communityMulti)
plot(marriageNet, vertex.size = 3, vertex.label = NA)
```

This becomes less feasible as the number of groups increases!!

## Modularity

Modularity takes a given group structure and calculates how separated the different groups are from each other. It therefore operationalizes our notion of groups above by calculating the proportion of ties that are within groups as opposed to between them. Networks with high modularity have dense connections between the nodes within modules but sparse connections between nodes in different modules.


We can think of modularity as both a measure of how effective a given grouping algorithm is - i.e. higher modularity means the alogrithm is identifying distinct, sociall separate groups. But it can also be thought of as a measure of the saliency of groups to the network in general. The higher modularity the more that groups structure the network. Modularity measures the extent to which a network can be divided into distinct modules or groups.

Getting a network's modularity in igraph is easy! 

```{r}
communityMulti <- multilevel.community(marriageNet)
communityMulti$modularity
```

We can either access the modularity score directly 
OR 

```{r}
modularity(communityWalk) # We can use the modularity() function on a group detection output.
```